Synthesizing Symmetric Lenses

### 2019-05-30 papers/SymmetricLenses/Conference/main.tex
update main and cover letter

\author{Solomon Maina}
\affiliation{
  \institution{Princeton University}
  \country{USA}
}
\email{smaina@cis.upenn.edu}

@@

\maketitle
%\fi

@@

\textit{Synthesizing} lenses can make the task even easier. Indeed, lens
languages like Boomerang are excellent targets for search-based synthesis,
since their 
specialized, sub-Turing-complete syntax and their expressive type systems
drastically limit the space of possible programs.
%
The recent Optician synthesizer for Boomerang~\cite{optician} has
demonstrated the feasibility of synthesizing quite complex examples, albeit
only for a restricted class of lenses---so-called bijective lenses.
Given source and target formats plus a small number of examples of
corresponding data, Optician's synthesis algorithm is guaranteed to find a
bijective lens matching the examples, if one exists, often requiring no
examples at all. While specifying regular expressions can be cumbersome, doing so
is much easier than programming lenses -- regular expressions are widely
understood whereas few programmers are fluent in lenses, and writing lenses
require users reason about how to align two formats at once.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis can be so effective, even relative to
successful synthesis tools in other domains, is that there are typically not
very many bijections between a given pair of data formats. 
If the synthesis algorithm finds any bijection, it is fairly likely to be
the intended one.

@@

    $\ProbabilityOf{\PRegexStar{\Regex}{\Probability}}{\String}$
    & =
    & $\Sigma_n \Sigma_{\String = \String_1 \ldots \String_n}\Probability^n(1-\Probability)\Pi_{i=1}^n\ProbabilityOf{\Regex}{\String_i}$\\
  \end{tabular}
\end{center}
One may think of SREs as string generators. Under this interpretation, the
constant SRE $\String$ always generates the string $\String$ and never any other
string. The SRE $\PRegexOr{\Regex_1}{\Regex_2}{\Probability}$ generates a string
from $\Regex_1$ with probability $\Probability$ and generates a string from
$\Regex_2$ with probability $1-\Probability$. The SRE
$\PRegexStar{\Regex}{\Probability}$ generates strings in $\Regex^n$ with
probability $\Probability^n(1-\Probability)$. For example, if the empty string
is not in $\Regex$, $\PRegexStar{\Regex}{\Probability}$ will generate the empty
string with probability $1-\Probability$.

@@

As an example, consider searching for a lens between \lstinline{"" | name.name$^*$}
and \lstinline{"" | name}. \GreedySynth might first the
sequence lens between the
sequences \lstinline{""} and \lstinline{""}, because it is a bijective
sequence lens between. As neither \lstinline{""} is involved in a sequence
lens, this lens is considered useful. Next, the sequence lens between
\lstinline{name.name$^*$} and \lstinline{name} would be popped: while that lens
is not bijective it is still better than the alternatives. As all sequences are
now involved in sequence lenses, and there are no examples to make other lenses
useful, no more sequence lenses would be added to the lens builder.

@@

Each task consists of a source and target regular expression and a set of
examples. We manually programmed the regular expressions. However, when writing
these regular expressions, we did not massage them to be amenable to synthesis.
The synthesizer implements the star-semiring regular expression equivalence
axioms, so that users do not have to worry about the way they write expressions
themselves. We are synthesizing transformations between real formats, so these
tasks are can be quite large -- on average, the size of each RE in the
specification is 1637 and the size of the synthesized lens is 3254.

@@

For each benchmark in the suite and each mode, we ran the system with a timeout
of 60 seconds, averaging the result over 5 runs. Figure~\ref{fig:times}
summarizes the results of these tests. We find that our algorithm is able to
synthesize all of the benchmarks in under 30 seconds. Without compositional
synthesis, the synthesis algorithm is able to solve 31 out of 48 problem instances.

@@

Both the bijective synthesis and the symmetric synthesis engines use a pair of
collaborating synthesizers that involves (1) a search for a compatible pair of
regular expressions and (2) search for a lens given those regular expressions.
Bijective synthesis is faster than symmetric synthesis because part (2) is much
faster. Specifically, a bijection must translate all data on the left into data
on the right, and this fact constrains the search. By contrast, a symmetric
synthesis problem has a choice of which data on the left to translate into data
on the right. This gives rise to a large set of possible choices, and symmetric
synthesis must consider all of them.

@@

Figure~\ref{fig:metric} summarizes the result of these experiments. The data
reveal that the information-theoretic metric is critical for finding the correct
lens: Only 10 of the benchmarks succeeded when running in \CCOpt{} mode. The
termination condition is also quite important. When running in \FLOpt{} mode,
the algorithm only discovers 5 lenses, which shows that the first class that
contains a satisfying lens is rarely the correct class. However, our algorithm
is not perfect and fails when either it is very difficult to find the desired
lens (necessitating \SRequire) or when a large amount of data is projected
(necessitating \Skip). Without any annotations, our algorithm finds the correct
lens for 40 of our 48 benchmarks; eight required relevance annotations to find
the correct lens.
% end evaluation

@@

More broadly, information theory and probabilistic languages are common tools in
natural language understanding, machine translation, information retrieval, data
extraction and grammatical inference (see \citet{pereira:info-theory}, for an
introduction). Indeed, our work was inspired, in part, by the PADS format
inference tool~\cite{pads:synthesis}, which was in turn inspired by earlier work
on probabilistic grammatical inference and information
extraction~\cite{kushmerick-thesis,arasu:extracting}. PADS did not synthesize
data transformers, and we are not aware of the use of information-theoretic
principles in type-directed or syntax-guided synthesis of DSL programs. More
recently, the principles used in \citet{pads:synthesis} have been applied in
unsupervised learning~\cite{ULPS}. These works learn compact descriptions of a
single data set. Ideally, such descriptions are compact and information theory
is used as a measure of the compactness of the description learned. In contrast,
we are attempting to learn a translation from data set to another. However,
there are many different candidate translations. To select amongst the candidate
translations, we choose the translation that preserves as much information from
source to target (and vice versa) as possible.

>>>>

\author{Solomon Maina}
\affiliation{
  \institution{Princeton University}
  \country{USA}
}
\email{smaina@seas.upenn.edu}

@@

\maketitle
\renewcommand{\shortauthors}{A. Miltner, S. Maina, K. Fisher, B. C. Pierce, D. Walker, S. Zdancewic}
%\fi

@@

\textit{Synthesizing} lenses can make the task even easier. Indeed, lens
languages like Boomerang are excellent targets for search-based synthesis,
since their 
specialized, sub-Turing-complete syntax and their expressive type systems
drastically limit the space of possible programs.
%
The recent Optician synthesizer for Boomerang~\cite{optician} has
demonstrated the feasibility of synthesizing quite complex examples, albeit
only for a restricted class of lenses---so-called bijective lenses.
Given source and target formats plus a small number of examples of
corresponding data, Optician's synthesis algorithm is guaranteed to find a
bijective lens matching the examples, if one exists, often requiring no
examples at all. While specifying regular expressions can be cumbersome, doing so
is much easier than programming lenses -- regular expressions are widely
understood whereas few programmers are fluent in lenses, and writing lenses
requires users to reason about how to align two formats at once.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis can be so effective, even relative to
successful synthesis tools in other domains, is that there are typically not
very many bijections between a given pair of data formats. 
If the synthesis algorithm finds any bijection, it is fairly likely to be
the intended one.

@@

    $\ProbabilityOf{\PRegexStar{\Regex}{\Probability}}{\String}$
    & =
    & $\Sigma_n \Sigma_{\String = \String_1 \ldots \String_n}\Probability^n(1-\Probability)\Pi_{i=1}^n\ProbabilityOf{\Regex}{\String_i}$\\
  \end{tabular}
\end{center}
One may think of SREs as string generators. Under this interpretation, the
constant SRE $\String$ always generates the string $\String$ and never any other
string. The SRE $\PRegexOr{\Regex_1}{\Regex_2}{\Probability}$ generates a string
from $\Regex_1$ with probability $\Probability$ and generates a string from
$\Regex_2$ with probability $1-\Probability$. The SRE
$\PRegexStar{\Regex}{\Probability}$ generates strings in $\Regex^n$ with
probability $\Probability^n(1-\Probability)$. For example,
$\PRegexStar{\Regex}{\Probability}$ will generate a string in $\Regex$ with
probability $\Probability(1-\Probability)$ and a string in
$\RegexConcat{\Regex}{\Regex}$ with probability $\Probability^2(1-\Probability)$.

@@

As an example, consider searching for a lens between \lstinline{"" | name.name$^*$}
and \lstinline{"" | name}. \GreedySynth might first pop the
sequence lens between the
sequences \lstinline{""} and \lstinline{""}, because it is a bijective
sequence lens between. As neither \lstinline{""} is involved in a sequence
lens, this lens is considered useful. Next, the sequence lens between
\lstinline{name.name$^*$} and \lstinline{name} would be popped: while that lens
is not bijective it is still better than the alternatives. As all sequences are
now involved in sequence lenses, and there are no examples to make other lenses
useful, no more sequence lenses would be added to the lens builder.

@@

Each task consists of a source and target regular expression and a set of
examples. We manually programmed the regular expressions. However, when writing
these regular expressions, we did not massage them to be amenable to synthesis.
The synthesizer implements the star-semiring regular expression equivalence
axioms, so that users do not have to worry about the way they write expressions
themselves. We are synthesizing transformations between real formats, so these
tasks are can be quite large -- on average, the size of each RE in the
specification is 1637 and the size of the synthesized lens is 3254. For each of
these tasks, we selected tasks until we did not feel additional examples would
be elucidating.

@@

For each benchmark in the suite and each mode, we ran the system with a timeout
of 60 seconds, averaging the result over 5 runs. Figure~\ref{fig:times}
summarizes the results of these tests. We find that our algorithm is able to
synthesize all of the benchmarks in under 30 seconds. Without compositional
synthesis, the synthesis algorithm is able to solve 31 out of 48 problem
instances. In total, 73 existing lenses were used in compositional synthesis
(about 1.5 per benchmark on average).

@@

Both the bijective synthesis and the symmetric synthesis engines use a pair of collaborating
synthesizers that (1) search for a compatible pair of regular expressions and (2) search
for a lens given those regular expressions. Bijective synthesis is faster than symmetric synthesis
because part (2) is much faster. Specifically, a bijection must translate all data on the left into data
on the right, and this fact constrains the search. By contrast, a symmetric synthesis problem has a
choice of which data on the left to translate into data on the right. This choice gives rise to a large set of
other choices, and symmetric synthesis must consider all of them.

@@

Figure~\ref{fig:metric} summarizes the result of these experiments. The data
reveal that the information-theoretic metric is critical for finding the correct
lens: Only 10 of the benchmarks succeeded when running in \CCOpt{} mode. The
termination condition is also quite important. When running in \FLOpt{} mode,
the algorithm only discovers 5 lenses, which shows that the first class that
contains a satisfying lens is rarely the correct class. However, our algorithm
is not perfect and fails when either it is very difficult to find the desired
lens (necessitating \SRequire) or when a large amount of data is projected
(necessitating \Skip). Without any annotations, our algorithm finds the correct
lens for 40 of our 48 benchmarks; eight required relevance annotations to find
the correct lens. In total, there are 12 uses of \SRequire, and 4 uses of \Skip
in our benchmark suite.
% end evaluation

@@

More broadly, information theory and probabilistic languages are common tools in
natural language understanding, machine translation, information retrieval, data
extraction and grammatical inference (see \citet{pereira:info-theory}, for an
introduction). Indeed, our work was inspired, in part, by the PADS format
inference tool~\cite{pads:synthesis}, which was in turn inspired by earlier work
on probabilistic grammatical inference and information
extraction~\cite{kushmerick-thesis,arasu:extracting}. PADS did not synthesize
data transformers, and we are not aware of the use of information-theoretic
principles in type-directed or syntax-guided synthesis of DSL programs. More
recently, the principles used in \citet{pads:synthesis} have been applied in
unsupervised learning~\cite{ULPS}. This work learns compact descriptions of a
single data set. Ideally, such descriptions are compact and information theory
is used as a measure of the compactness of the description learned. In contrast,
we are attempting to learn a translation from one data set to another. However,
there are many different candidate translations. To select amongst the candidate
translations, we choose the translation that preserves as much information from
source to target (and vice versa) as possible.
